Amphistomy increases leaf photosynthesis more in coastal than montane plants of Hawaiian ʻilima (*Sida fallax*)

### 2023-10-26 ms/ms.qmd
revising ms

$^2$ School of Life Sciences, University of California, Davis, CA, USA 95616

@@

\textbf{Key results}  Coastal ʻilima leaves benefit `r prettyNum(habitat_aa$times_increase, digits = 3L)` times more from amphistomy compared to their montane counterparts. Our evidence was equivocal with respect to two hypotheses – that coastal leaves benefit more because they are thicker and therefore have lower CO$_2$ conductance through the internal airspace, and that leaves with more similar conductances at each surface benefit more than leaves with few stomata on the upper surface.

@@

**Keywords:** amphstomy, leaf, light, Hawaiʻi, leaf, *Sida fallax*, stomata

@@

Amphistomy, the presence of stomata on both lower and upper surfaces of broad leaves, should increase carbon gain by reducing the average diffusion pathlength between stomata and chloroplasts, yet paradoxically this seemingly simple adaptation is uncommon in nature and we don't know why. Understanding variation in stomatal traits like amphistomy is imperative because these tiny pores play an outsized ecological role in the global carbon and water cycles [@hetherington_role_2003; @berry_stomata:_2010]. A widely applicable, accurate representation of how stomata mediate the relationship between CO$_2$ gained through photosynthesis and water lost through evapotranspiration is essential to predict future climate using Earth Systems Models [@jarvis_interpretation_1976; @biggins_model_1987; @collatz_physiological_1991; @leuning_critical_1995; @sellers_modeling_1997]. Optimality models accurately predict the major cause of water loss, stomatal conductance ($g_\mathrm{sw}$), by assuming plants maximize carbon gain minus a cost of water [@cowan_stomatal_1977; @givnish_economy_1986; @medlyn_reconciling_2011; @lin_optimal_2015; @wang_towards_2017; @franks_comparing_2018; @deans_optimization_2020; @franklin_organizing_2020; @wang_theoretical_2020; @harrison_ecoevolutionary_2021]. Despite the success of optimality modeling in predicting $g_\mathrm{sw}$, the same modeling approach has so far failed to explain the rarity of amphistomatous leaves [@muir_is_2019]. **This gap between theory and observations strongly implies that we remain ignorant about some key benefits and costs associated with stomata.**

@@

Amphistomy should also enhance photosynthesis when leaf boundary layer resistance is high, because apportioning total flux between two boundary layers rather than one results in a smaller CO$_2$ concentration drop between the atmosphere and stomata. A similar effect has been validated with a computer model and measurements for transpiration: amphistomatous leaves lose somewhat more water for the same vapor pressure deficit and total $g_\mathrm{sw}$ [@foster_influence_1986], but the additional carbon gain should be enough to offset this cost under most realistic conditions [@muir_is_2019].

@@

```{r, locations, echo=FALSE,message=FALSE,warning=FALSE}
#| label: tbl-locations
#| tbl-cap: "ʻIlima study sites for this study."
#| 
read_rds("../processed-data/site.rds") |>
  filter(site_code != "nnpl") |>
  mutate(Island = case_when(
    island == "oahu" ~ "Oʻahu",
    island == "hawaii" ~ "Hawaiʻi"
  )) |>
  arrange(site_type, site) |>
  select(Site = site, Island, Habitat = site_type, Latitude = latitude_degree,
         Longitude = longitude_degree, `Elevation (mas)` = elevation_m) |>
  knitr::kable(format = "latex", digits = c(rep(0, 3), 3, 3, 0),
               booktabs = TRUE) |>
  kableExtra::kable_styling(latex_options = c("striped", "no_vertical_lines")) |>
  kableExtra::column_spec(1, width = "4cm") |>
  kableExtra::column_spec(2, width = "1.5cm") |>
  kableExtra::column_spec(3, width = "1.5cm") |>
  kableExtra::column_spec(4, width = "1.5cm") |>
  kableExtra::column_spec(5, width = "1.5cm") |>
  kableExtra::column_spec(6, width = "1.5cm")

@@

The log-response ratio is easy to interpret because values above 0 indicate a photosynthetic advantage of amphistomy, whereas values less than 0 indicate a disadvantage. The response ratio is also better than the absolute difference because it indicates a proportional change in $A$, which facilitates comparisons across leaves and environments with different baseline photosynthetic rates. The irradiance of the light source in the pseudohypo leaf is higher because the propafilm reduces transmission. Because the propafilm reduces transmission, we used a higher incident $\mathrm{PPFD}$ for pseudohypo leaves (higher by a factor 1/0.91, the inverse of the measured transmissivity of the propafilm). We also set the stomatal conductance ratio, for purposes of calculating boundary layer conductance, to 0 for pseudohypo leaves following manufacturer directions.

To control for order effects, we alternated between starting with amphi (untreated) or pseudohypo (treated) leaf measurements. In the field, we acclimated the focal leaf to high light ($\mathrm{PPFD} = 2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$), high relative humidity ($\mathrm{RH} = 70\%$), $T_\text{leaf} = 25-30~° \text{C}$, depending on ambient temperature, until $A$ and $g_\text{sw}$ reach their maximum. We used these data as our estimates of maximum $A$ and $g_\text{sw}$. After that, we decreased $\mathrm{RH}$ to $\approx 10\%$ to induce rapid stomatal closure without photosynthetic downregulation. All other environmental conditions in the leaf chamber remained the same. We logged data until $g_\text{sw}$ reached its nadir. We then repeated the process with the other treatment (amphi or pseudohypo).

@@

It is not feasible to record $A_\mathrm{amphi}$ and $A_\mathrm{hypo}$ at the exact same $g_\text{sw}$. To overcome this, we fit $A \textendash g_\text{sw}$ curves using a linear regression of $\text{log}(g_\mathrm{sw})$ on $A$ to interpolate modeled $A$ for amphi and pseudohypo leaves at the same $g_\text{sw}$. In 10 of 12 leaves, the minimum $g_\text{sw}$ of the amphi curve was smaller than the maximum $g_\text{sw}$ of the pseudohypo curve (i.e. the curves overlapped for a range of $g_\text{sw}$ values). In those cases, we estimated $\mathrm{AA}$ at the $g_\mathrm{sw}$ value in the middle of the range of overlap. In 2 of 12 leaves, the $A \textendash g_\text{sw}$ curves did not quite overlap because the minimum $g_\text{sw}$ of the amphi curve was slightly greater than the maximum $g_\text{sw}$ of the pseudohypo curve. In those cases, we estimated $\mathrm{AA}$ by extrapolating slightly, `r gsw_extrap`, beyond the measured curves to the $g_\mathrm{sw}$ value in between the curves. We fit Bayesian regressions using the *R* package **brms** version `r fit_licor$version$brms` [@burkner_brms_2017] with MCMC sampling in *Stan* [@stan_development_team_stan_2023]. We used CmdStan version `r fit_licor$version$cmdstan` and **cmdstanr** version `r fit_licor$version$cmdstanr` [@gabry_cmdstanr_2023] to interface with *R* version `r getRversion()` [@r_core_team_r:_2023]. We sampled the posterior distribution from `r dim_fit_licor[2]` chains with `r dim_fit_licor[1]` iterations each after `r dim_fit_licor[1]` warmup iterations. We estimated parameters and confidence intervals as the median and 95% quantile intervals of the posterior, respectively. The key prediction is that $\mathrm{AA}_\text{coastal} > \mathrm{AA}_\text{montane}$.

@@

We tested whether leaf thickness (log-transformed) varied between coastal and montane populations and among individuals within populations using a Bayesian mixed-effects model with habitat as a fixed effect and individual plant and site as random effects. We used the *R* package **brms** version `r fit_thickness$version$brms` [@burkner_brms_2017] to fit the model in *Stan* [@stan_development_team_stan_2023] with CmdStan version `r fit_thickness$version$cmdstan` and **cmdstanr** version `r fit_thickness$version$cmdstanr` [@gabry_cmdstanr_2023]. We sampled the posterior distribution from `r dim_fit_thickness[2]` chains with `r dim_fit_thickness[1]` iterations each after `r dim_fit_thickness[1]` warmup iterations. We estimated the relationship between population average leaf thickness and $\mathrm{AA}$ measured from a single individual per population. We used this approach most of the variation in leaf thickness was among sites and the plant selected for gas exchange measurements was not always among the plants randomly selected for leaf thickness, precluding individual level correlation. We propagated uncertainty about in $\mathrm{AA}$ and leaf thickness by integrating over the entire posterior distribution sample for each variable. The key prediction is that the effect of leaf thickness on $\mathrm{AA}$ is positive.

@@

Amphistomy is a seemingly simple way that leaves can increase carbon gain without significant additional water loss, yet it is rare in nature and we do not know why. The strong association between amphiostomy and sunny, open habitats suggests that amphistomy may benefit sun leaves more than shade leaves, but progress has been limited by the lack of evidence that amphiostomy *per se* affects improves photosynthesis in a given leaf. By experimentally blocking gas exchange through the upper surface in a controlled environment, we directly compared an amphistomatous leaf to an otherwise identical pseudohypostomatous leaf. This allows us to quantity the amphistomy advantage ($\mathrm{AA}$) holding all else constant. Taking advantage of the of the steep climatic gradients in the Hawaiian archipelago, we applied this new method to show for the first time that sun leaves benefit  `r prettyNum(habitat_aa$times_increase, digits = 3L)` times more from amphistomy than shade leaves on ʻilima (*Sida fallax*) plants ($\mathrm{AA}_\text{coastal} = `r prettyNum(habitat_aa[['coastal']], digits = 2L)`$ vs. $\mathrm{AA}_\text{montane} = `r prettyNum(habitat_aa[['montane']], digits = 2L)`$). While this result has not yet been validated in other species, our results indicate that part of the reason amphistomatous leaves are found most commonly in high light habitats is that the adaptive benefit is greater in this environment.

If $\mathrm{AA}$ is typically greater in sun leaves than shade leaves, it could partially explain the distribution of amphi- and hypostoamtous leaves, but the precise mechanism(s) require further study. One hypothesis is that the internal airspace conductance, $g_\mathrm{ias}$, from stomata to mesophyll cell walls is lower in thicker sun leaves [@parkhurst_adaptive_1978]. All else being equal, a leaf with lower $g_\mathrm{ias}$ will benefit more from amphistomy. Our results partially support this hypothesis. Coastal ʻilima leaves with high $\mathrm{AA}$ (\autoref{fig:habitat-aa}) are thicker than montane leaves, but the relationship between $\mathrm{AA}$ and leaf thickness within habitats is actually slightly negative (\autoref{fig:traits-aa}A), opposite our prediction. Since coastal and montane leaves differ in many respects besides thickness, we do not have enough data to conclude that leaf thickness explains the variation in $\mathrm{AA}$ between habitats. Alternatively, other biochemical or anatomical differences between coastal and montane leaves may explain why $\mathrm{AA}$ is greater in coastal leaves. The negative relationship between leaf thickness and $\mathrm{AA}$ could be explained if thicker leaves compensated by having a more porous mesophyll and/or less tortuous airspaces [@theroux-rancourt_maximum_2021].

@@

This study reports the first direct experimental evidence that having stomata open on both leaf surfaces, amphistomy, increases photosynthesis for a given total stomatal conductance, particularly in leaves from the type of open, sunny habitats where this trait is most common. By developing a straightforward gas exchange method to block gas exchange through the upper surface, we directly compared the photosynthetic rate of a leaf with gas exchange through both surfaces or just one, holding all other factors constant. In doing so, we found that coastal leaves of the indigenous Hawaiian ʻilima (*Sida fallax*) enjoyed a greater photosynthetic benefit from amphistomy than nearby montane leaves living in more closed forest. We cannot yet ascribe the difference in amphistomy advantage between coastal and montane leaves to particular physiological or anatomical variation, but this is a promising area for future research.

>>>>

$^2$ Department of Plant Sciences, University of California, Davis, CA, USA 95616

@@

\textbf{Key results}  Coastal ʻilima leaves benefit `r prettyNum(habitat_aa$times_increase, digits = 3L)` times more from amphistomy compared to their montane counterparts. Our evidence was equivocal with respect to two hypotheses – that coastal leaves benefit more because 1) they are thicker and therefore have lower CO$_2$ conductance through the internal airspace, and 2) that they benefit more because they have similar conductance on each surface, as opposed to most of the conductance being on the lower (abaxial) surface.

@@

**Keywords:** amphstomy, leaf, light, Hawaiʻi, *Sida fallax*, stomata

@@

Amphistomy, the presence of stomata on both lower and upper surfaces of broad leaves, should increase carbon gain by reducing the average diffusion pathlength between stomata and chloroplasts, yet paradoxically this seemingly simple adaptation is uncommon in nature and we don't know why. Understanding variation in stomatal traits like amphistomy is imperative because these tiny pores play an outsized ecological role in the global carbon and water cycles [@hetherington_role_2003; @berry_stomata:_2010]. A widely applicable, accurate representation of how stomata mediate the relationship between CO$_2$ gained through photosynthesis and water lost through transpiration is essential to predict future climate using Earth Systems Models [@jarvis_interpretation_1976; @biggins_model_1987; @collatz_physiological_1991; @leuning_critical_1995; @sellers_modeling_1997]. Optimality models accurately predict the major cause of water loss, stomatal conductance ($g_\mathrm{sw}$), by assuming plants maximize carbon gain minus a cost of water [@cowan_stomatal_1977; @givnish_economy_1986; @medlyn_reconciling_2011; @lin_optimal_2015; @wang_towards_2017; @franks_comparing_2018; @deans_optimization_2020; @franklin_organizing_2020; @wang_theoretical_2020; @harrison_ecoevolutionary_2021]. Despite the success of optimality modeling in predicting $g_\mathrm{sw}$, the same modeling approach has so far failed to explain the rarity of amphistomatous leaves [@muir_is_2019]. **This gap between theory and observations strongly implies that we remain ignorant about some key benefits and costs associated with stomata.**

@@

Amphistomy should also enhance photosynthesis when leaf boundary layer resistance is high, because apportioning total flux between two boundary layers rather than one results in a smaller CO$_2$ concentration drop between the atmosphere and stomata. A similar effect has been validated with a computer model and measurements for transpiration: amphistomatous leaves lose somewhat more water for the same vapor pressure deficit and total $g_\mathrm{sw}$ [@foster_influence_1986], but the additional carbon gain should be enough to offset this cost under most realistic conditions [@muir_is_2019]. However, if minimal stomatal conductance is related to stomatal density [@drake_smaller_2013; @marquez_cuticular_2021] and the upper boundary layer conductance is higher, then amphistomy could cause additional, unavoidable water loss.

@@

```{r, locations, echo=FALSE,message=FALSE,warning=FALSE}
#| label: tbl-locations
#| tbl-cap: "ʻIlima study site location information."
#| 
read_rds("../processed-data/site.rds") |>
  filter(site_code != "nnpl") |>
  mutate(Island = case_when(
    island == "oahu" ~ "Oʻahu",
    island == "hawaii" ~ "Hawaiʻi"
  )) |>
  arrange(site_type, site) |>
  select(Site = site, Island, Habitat = site_type, Latitude = latitude_degree,
         Longitude = longitude_degree, `Elevation (mas)` = elevation_m) |>
  knitr::kable(format = "latex", digits = c(rep(0, 3), 3, 3, 0),
               booktabs = TRUE) |>
  kableExtra::kable_styling(latex_options = c("striped", "no_vertical_lines")) |>
  kableExtra::column_spec(1, width = "4cm") |>
  kableExtra::column_spec(2, width = "1.5cm") |>
  kableExtra::column_spec(3, width = "1.5cm") |>
  kableExtra::column_spec(4, width = "1.5cm") |>
  kableExtra::column_spec(5, width = "1.5cm") |>
  kableExtra::column_spec(6, width = "1.5cm")

@@

The log-response ratio is straightforward to interpret because values above 0 indicate a photosynthetic advantage of amphistomy, whereas values less than 0 indicate a disadvantage. The response ratio is also better than the absolute difference because it indicates a proportional change in $A$, which facilitates comparisons across leaves and environments with different baseline photosynthetic rates. The irradiance of the light source in the pseudohypo leaf is higher because the propafilm reduces transmission. To compensate for reduced transmission, we increased incident $\mathrm{PPFD}$ for pseudohypo leaves by a factor 1/0.91, the inverse of the measured transmissivity of the propafilm. We also set the stomatal conductance ratio, for purposes of calculating boundary layer conductance, to 0 for pseudohypo leaves following manufacturer directions.

To control for order effects, we alternated between starting with amphi (untreated) or pseudohypo (treated) leaf measurements. In the field, we acclimated the focal leaf to high light ($\mathrm{PPFD} = 2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$), high relative humidity ($\mathrm{RH} = 70\%$), $T_\text{leaf} = 25-30~° \text{C}$, depending on ambient temperature, until $A$ and $g_\text{sw}$ reach their maximum. We used these data as our estimates of maximum $A$ and $g_\text{sw}$. After that, we decreased $\mathrm{RH}$ to $\approx 10\%$ to induce rapid stomatal closure without photosynthetic downregulation. Hence, $A_\text{amphi}$ and $A_\text{hypo}$ were both measured at low chamber humidity after the leaf had acclimated to high humidity. All other environmental conditions in the leaf chamber remained the same. We logged data until $g_\text{sw}$ reached its nadir. We then repeated the process with the other treatment (amphi or pseudohypo).

@@

It is not feasible to record $A_\mathrm{amphi}$ and $A_\mathrm{hypo}$ at the exact same $g_\text{sw}$. To overcome this, we fit $A \textendash g_\text{sw}$ curves using a linear regression of $\text{log}(g_\mathrm{sw})$ on $A$ to interpolate modeled $A$ for amphi and pseudohypo leaves at the same $g_\text{sw}$. In 10 of 12 leaves, the minimum $g_\text{sw}$ of the amphi curve was smaller than the maximum $g_\text{sw}$ of the pseudohypo curve (i.e. the curves overlapped for a range of $g_\text{sw}$ values). In those cases, we estimated $\mathrm{AA}$ at the $g_\mathrm{sw}$ value in the middle of the range of overlap. In 2 of 12 leaves, the $A \textendash g_\text{sw}$ curves did not quite overlap because the minimum $g_\text{sw}$ of the amphi curve was slightly greater than the maximum $g_\text{sw}$ of the pseudohypo curve. In those cases, we estimated $\mathrm{AA}$ by extrapolating slightly, `r gsw_extrap`, beyond the measured curves to the $g_\mathrm{sw}$ value in between the curves. We fit Bayesian regressions using the *R* package **brms** version `r fit_licor$version$brms` [@burkner_brms_2017] with MCMC sampling in *Stan* [@stan_development_team_stan_2023]. We used CmdStan version `r fit_licor$version$cmdstan` and **cmdstanr** version `r fit_licor$version$cmdstanr` [@gabry_cmdstanr_2023] to interface with *R* version `r getRversion()` [@r_core_team_r:_2023]. We sampled the posterior distribution from `r dim_fit_licor[2]` chains with `r dim_fit_licor[1]` iterations each after `r dim_fit_licor[1]` warmup iterations. We estimated parameters and confidence intervals as the median and 95% quantile intervals of the posterior, respectively. The key prediction is that $\mathrm{AA}_\text{coastal} > \mathrm{AA}_\text{montane}$, meaning the 95\% confidence intervals of $\mathrm{AA}_\text{coastal} - \mathrm{AA}_\text{montane}$ should be positive and not encompass 0.

@@

We tested whether leaf thickness (log-transformed) varied between coastal and montane populations and among individuals within populations using a Bayesian mixed-effects model with habitat as a fixed effect and individual plant and site as random effects. We used the *R* package **brms** version `r fit_thickness$version$brms` [@burkner_brms_2017] to fit the model in *Stan* [@stan_development_team_stan_2023] with CmdStan version `r fit_thickness$version$cmdstan` and **cmdstanr** version `r fit_thickness$version$cmdstanr` [@gabry_cmdstanr_2023]. We sampled the posterior distribution from `r dim_fit_thickness[2]` chains with `r dim_fit_thickness[1]` iterations each after `r dim_fit_thickness[1]` warmup iterations. We estimated the relationship between population average leaf thickness and $\mathrm{AA}$ measured from a single individual per population. We used this approach most of the variation in leaf thickness was among sites and the plant selected for gas exchange measurements was not always among the plants randomly selected for leaf thickness, precluding individual level correlation. We propagated uncertainty about in $\mathrm{AA}$ and leaf thickness by integrating over the entire posterior distribution sample for each variable. The key prediction is that the effect of leaf thickness on $\mathrm{AA}$ is positive, meaning the 95\% confidence interval of the slope should be positive and not encompass 0.

@@

Amphistomy is a seemingly simple way that leaves can increase carbon gain without significant additional water loss, yet it is rare in nature and we do not know why. The strong association between amphiostomy and sunny, open habitats suggests that amphistomy may benefit sun leaves more than shade leaves, but progress has been limited by the lack of evidence that amphiostomy *per se* affects improves photosynthesis in a given leaf. By experimentally blocking gas exchange through the upper surface in a controlled environment, we directly compared an amphistomatous leaf to an otherwise identical pseudohypostomatous leaf. This allows us to quantity the amphistomy advantage ($\mathrm{AA}$) holding all else constant. Taking advantage of the steep climatic gradients in the Hawaiian archipelago, we applied this new method to show for the first time that sun leaves benefit  `r prettyNum(habitat_aa$times_increase, digits = 3L)` times more from amphistomy than shade leaves on ʻilima (*Sida fallax*) plants ($\mathrm{AA}_\text{coastal} = `r prettyNum(habitat_aa[['coastal']], digits = 2L)`$ vs. $\mathrm{AA}_\text{montane} = `r prettyNum(habitat_aa[['montane']], digits = 2L)`$). While this result has not yet been validated in other species, our results indicate that part of the reason amphistomatous leaves are found most commonly in high light habitats is that the adaptive benefit is greater in this environment.

If $\mathrm{AA}$ is typically greater in sun leaves than shade leaves, it could partially explain the distribution of amphi- and hypostoamtous leaves, but the precise mechanism(s) require further study. One hypothesis is that the internal airspace conductance, $g_\mathrm{ias}$, from stomata to mesophyll cell walls is lower in thicker sun leaves [@parkhurst_adaptive_1978]. All else being equal, a leaf with lower $g_\mathrm{ias}$ will benefit more from amphistomy. Our results partially support this hypothesis. Coastal ʻilima leaves with high $\mathrm{AA}$ (\autoref{fig:habitat-aa}) are thicker than montane leaves, but the relationship between $\mathrm{AA}$ and leaf thickness within habitats is actually slightly negative (\autoref{fig:traits-aa}A), opposite our prediction. Since coastal and montane leaves differ in many respects besides thickness, we do not have enough data to conclude that leaf thickness explains the variation in $\mathrm{AA}$ between habitats. Alternatively, other biochemical or anatomical differences between coastal and montane leaves may explain why $\mathrm{AA}$ is greater in coastal leaves. The negative relationship, (albeit nonsignificant in that 95\% confidence intervals encompassed 0), between leaf thickness and $\mathrm{AA}$ could be explained if thicker leaves compensated by having a more porous mesophyll and/or less tortuous airspaces [@theroux-rancourt_maximum_2021].

@@

This study reports the first direct experimental evidence that having stomata open on both leaf surfaces, amphistomy, increases photosynthesis for a given total stomatal conductance, particularly in leaves from the type of open, sunny habitats where this trait is most common. By developing a straightforward experimental method to block gas exchange through the upper surface, we directly compared the photosynthetic rate of a leaf with gas exchange through both surfaces or just one, holding all other factors constant. In doing so, we found that coastal leaves of the indigenous Hawaiian ʻilima (*Sida fallax*) enjoyed a greater photosynthetic benefit from amphistomy than nearby montane leaves living in more closed forest. This is not because amphistomatous leaves necessarily have greater leaf surface available for stomata, although that be an important consideration in predicting realized photosynthetic rates in nature. Rather, our experiments show that coastal amphistomatous leaves with the same total leaf stomatal conductance photosynthesize more than identical hypostomatous leaves. We cannot yet ascribe the difference in amphistomy advantage between coastal and montane leaves to particular physiological or anatomical variation, but this is a promising area for future research.
